\documentclass{beamer}
\usetheme{focus}

\usefonttheme{serif}

\usepackage{fontspec}
%\setmainfont[Ligatures=TeX]{DejaVu Sans}
\setmainfont{Charis SIL}

\usepackage{pythontex}

\newcommand{\slide}[4]{
\begin{frame}
	\begin{center}
		\vfill
		
		\textbf{\huge #1}
		
		\vfill
		
		{\large #2}
		
		\vfill
		
		{\huge #3}
		
		\vfill
		
		\emph{#4}
		
	\end{center}
\end{frame}
}

\newcommand{\interlinearslide}[3]{
\begin{frame}
	\begin{center}
		\vfill
		
		\textbf{\huge #1}
		
		\vfill
		
		{\large #2}
		
		\vfill
		
		{\huge #3}
		
		
	\end{center}
\end{frame}
}

\begin{document}

\begin{frame}

\end{frame}


\interlinearslide{Ʇaula Nǃòõ Uǃqʼa maã Jùu aʇe ǂu ełǁxa iñi}{ ǀaula ᵑǃɔ̰̃ː uǃ͡qʼa mãː ɟṵː aǀe ǂu eɬǁ͡χa iɲi}{Once upon a time, the North Wind and the Sun were discussing}

\interlinearslide{nʇum nui ʼèe sʇaʇau ǁùuṇa.}{ ᵑǀmː n̪ui ʔḛː s̪ǀaǀau ǁṵːɳa.}{over which one of them two was stronger.}

\interlinearslide{Ṭurra, loõṇi ʇxoi onʇʼa oǃʼo ṣǂaañi.}{ ʈurra, lɔ̃ːɳi ǀ͡χɔi ɔ̃ᵑǀˀa ɔǃˀɔ ʂǂaːɲi.}{Suddenly, a vagabond wrapped in a warm cloak arrived to them.}

\interlinearslide{Maã aʇe ǂu nǃʼoirre nui ʼèe ǂa ʼa ṣǂaañi ku nǁxòi upa ʇxoi ṭèe,}{ mãː aǀe ǂu ᵑǃˀɔirre n̪ui ʔḛː ǂa ʔa ʂǂaːɲi ku ᵑǁ͡ʁɔ̰ḭ upa ǀ͡χɔi ʈḛː,}{The Wind and the Sun decided that the first of them that would make the vagabond take off the cloak,}

\interlinearslide{unʇu sʇaʇau tłìi šǃu nui.}{ mᵑǀu s̪ǀaǀau t͡ɬḭː šǃu n̪ui.}{truly that one would have been stronger.}

\interlinearslide{Nǃòõ Uǃqʼa maã ǂxàuʼa tṣe ra tsùu nʘaã,}{ ᵑǃɔ̰̃ː uǃ͡qʼa mãː ǂ͡χa̰ṵʔa ʈ͡ʂe ra t͡s̪ṵː ᵑʘãː,}{The North Wind blew as strong as he could,}

\interlinearslide{ṇam ča uji tsùu maã, ṣǂaañi ku ʼa ṭèe onʇʼa ǃqʼati nǃxùu.}{ ɳãm t͡ʃa uɟi t͡s̪ṵː mãː, ʂǂaːɲi ku ʔa ʈḛː ɔ̃ᵑǀˀa ǃ͡qʼat͡s̪i ᵑǃ͡ʁṵː.}{but as the wind blew, the vagabond enveloped themselves in the cloak ever more.}

\interlinearslide{ʼai Uǃqʼa maã eña ṣǂqʼo.}{ ʔai uǃ͡qʼa mãː eɲa ʂǂ͡qʼɔ.}{And so, the Wind gave up.}

\interlinearslide{ʼai ʼai, loõṇi ra čìiči Jùu aʇe, ʼai ǂu ku uǃqʼo ʇxoi ṭèe.}{ ʔai ʔai, lɔ̃ːɳi ra t͡ʃḭːt͡ʃi ɟṵː aǀe, ʔai ǂu ku uǃ͡qʼɔ ǀ͡χɔi ʈḛː.}{But then, the Sun shone warmly, and so the vagabond took off their cloak.}

\interlinearslide{Nǃòõ Uǃqʼa maã łǁa sʇaʇau Jùu aʇe.}{ ᵑǃɔ̰̃ː uǃ͡qʼa mãː ɬǁa s̪ǀaǀau ɟṵː aǀe.}{Thus, the North Wind saw that the Sun was stronger.}

%
%
%\slide{ǃòo}{/ǃɔ̰ː/}{1}{also: alone, unaccompained}
%\slide{nʇum}{/ᵑǀmː/}{2}{}
%\slide{eǂaaka}{/eǂaːka/}{3}{}
%\slide{sʇʼe}{/sǀ̪ˀe/}{4}{}
%\slide{šǃqʼoi}{šǃ͡qʼɔi}{5}{}
%\slide{aǁum}{aǁmː}{6}{}
%\slide{nǂoiči}{ᵑǂɔit͡ʃi}{7}{}
%\slide{noti}{/n̪ɔt͡si̪ /}{8}{}
%\slide{nʇaati}{...}{9}{}
%\slide{aǃùuma}{}{10}{}
%\slide{ŋum ʼa ǃòo}{}{11}{}
%\slide{ŋum}{}{12}{}
%\slide{ŋum ǂa}{}{13}{}
%\slide{ŋum ǃaaru}{}{14}{}
%\slide{šǃoǃqʼoi}{}{15}{}
%\slide{nonoti}{}{16}{}
%\slide{nonoti ǂa}{}{17}{}
%\slide{aǁumňu}{}{18}{}
%\slide{*pèe ʼa ǃòo}{}{19}{(rare form)}
%\slide{pèe}{}{20}{}
%\slide{*aǃuǃùuma}{}{20}{(rare alternate form)}
%\slide{pèe ǂa}{}{21}{}
%\slide{*pèe ǃaaru}{}{22}{(rare form)}
%\slide{*ŋuŋum ʼa ǃòo}{}{23}{(rare form)}
%\slide{ŋuŋum}{}{24}{}

\end{document}