\lettersection{ǁ}\dictentry{aǁum}{aǁmː}{\dictsense{card.\-num.}{six}{}}
\dictentry{ǁa}{ǁa}{\dictsense{card.\-num.}{few}{}}
\dictentry{ǁarra}{ǁarra}{\dictsense{v.\-intr.}{close one's eyes, not see, refuse to look}{}\dictsensesep\dictsense{v.\-intr.}{(in SVC) refuse to ..., not intend to ..., resist ...}{}}
\dictentry{ǁaũpe}{ǁãmpe}{\dictsense{n.}{foot}{\dictclassifiers{\dictref{ʇṵu}}}\dictsensesep\dictsense{v.\-intr.}{walk}{}}
\dictentry{ǁa̰a}{ǁa̰ː}{\dictsense{v.\-tr.}{hold}{}\dictsensesep\dictsense{clf.}{young women, recent mothers}{}}
\dictentry{ǁo̰i}{ǁɔ̰ḭ}{\dictsense{Postverb}{NEG, not, negates the action of the verb.}{\dictexample{Uji iǃo̰orri ǁo̰i ñḛʼe}{The old man doesn't eat anything.}}}
\dictentry{ǁṵuṇa}{ǁṵːɳa}{\dictsense{post.}{about}{}}
\lettersection{ǁʼ}\dictentry{ǁʼu}{ǁˀu}{\dictsense{Ditransitive verb}{give, provide, (A) give (T) to (P)}{}}
\lettersection{ǁx}\dictentry{ǁxa}{ǁ͡χa}{\dictsense{n.}{penis}{\dictclassifiers{\dictref{ʇṵu}}}}
\lettersection{ǁqʼ}\dictentry{ǁqʼa}{ǁ͡qʼa}{\dictsense{adj.}{big, large}{}}
\dictentry{ǁqʼooña}{ǁ͡qʼɔːɲa}{\dictsense{n.}{crab}{}\dictsensesep\dictsense{n.}{lobster}{}}
\dictentry{ǁqʼoṯe}{ǁ͡qʼɔt͡s̪e}{\dictsense{n.}{game, animals to be hunted, especially larger herbivores}{\dictclassifiers{\dictref{ǂʼa}}}}
\lettersection{ǁʛ}\dictentry{ǁʛa}{ǁ͡ʛa}{\dictsense{v.\-tr.}{scare, (P) be afraid of (A),}{}\dictsensesep\dictsense{n.}{fear}{\dictclassifiers{\dictref{ʇa}}}}
\lettersection{ɴǁ}\dictentry{iɴǁa̰a}{ĩᵑǁa̰ː}{\dictsense{v.\-intr.}{sleep}{}}
\dictentry{iɴǁoi}{ĩᵑǁɔi}{\dictsense{v.}{say}{}}
\dictentry{ɴǁañe}{ᵑǁaɲe}{\dictsense{Ditransitive verb}{(A) tell, recount, say (T) to (P)}{}}
\dictentry{ɴǁa̰ʼa}{ᵑǁa̰ʔa}{\dictsense{n.}{gold}{}}
\dictentry{ɴǁoi}{ᵑǁɔi}{\dictsense{v.}{can}{}}
\dictentry{ɴǁo̰õ}{ᵑǁɔ̰̃ː}{\dictsense{v.}{descend}{}}
\dictentry{ɴǁoṭro}{ᵑǁɔʈ͡ʂɔ}{\dictsense{n.}{door}{}}
\dictentry{ɴǁui}{ᵑǁui}{\dictsense{v.\-intr.}{jump}{}}
\lettersection{ɴǁʼ}\dictentry{ɴǁʼam}{ᵑǁˀãm}{\dictsense{n.}{story, tale}{\dictclassifiers{\dictref{ṇṵĩ}}}}
\lettersection{ɴǁx}\dictentry{ɴǁxape}{ᵑǁ͡ʁape}{\dictsense{v.\-tr.}{insult, berate, scold}{}}
\dictentry{ɴǁxo̰i}{ᵑǁ͡ʁɔ̰ḭ}{\dictsense{v.}{remove}{}}
\lettersection{ɬǁ}\dictentry{eɬǁa}{eɬǁa}{\dictsense{n.}{conflict, discussion, disagreement, verbal fight}{}}
\dictentry{iɬǁui}{iɬǁui}{\dictsense{n.}{milk}{\dictclassifiers{\dictref{ǂṵm}}}}
\dictentry{ɬǁa}{ɬǁa}{\dictsense{v.\-tr.}{see}{}}
\dictentry{ɬǁauʼi}{ɬǁauʔi}{\dictsense{adj.}{graceful, delicately beautiful}{}}
\dictentry{ɬǁoi}{ɬǁɔi}{\dictsense{clf.}{static formations, groups and sequences of inanimate objects, layouts, patterns and textures}{}}
\lettersection{ɬǁʼ}\dictentry{aɬǁʼi}{aɬǁˀi}{\dictsense{n.}{money}{}}
\dictentry{ɬǁʼa}{ɬǁˀa}{\dictsense{v.\-intr.}{transform}{}}
\lettersection{ɬǁqʼ}\dictentry{ɬǁqʼa}{ɬǁ͡qʼa}{\dictsense{n.}{egg}{\dictclassifiers{\dictref{ṉoõ}}}}
\lettersection{ǂ}\dictentry{eǂaaka}{eǂaːka}{\dictsense{card.\-num.}{three}{}}
\dictentry{uǂuuña}{uǂuːɲa}{\dictsense{n.}{song}{\dictclassifiers{\dictref{ʘa̰a}}}}
\dictentry{ǂa}{ǂa}{\dictsense{ord.\-num.}{first}{}}
\dictentry{ǂaã}{ǂãː}{\dictsense{v.\-intr.}{go, travel, move}{}\dictsensesep\dictsense{v.\-intr.}{(in SVC) begin ..., go to do..., go there and ...,}{}}
\dictentry{ǂaãñi}{ǂãːɲi}{\dictsense{v.}{flee}{}}
\dictentry{ǂaǂxa̰a}{ǂaǂ͡χa̰ː}{\dictsense{v.\-tr.}{snap, break (especially crack) in half}{}}
\dictentry{ǂa̰m}{ǂã̰m̰}{\dictsense{n.}{man}{}}
\dictentry{ǂoipe}{ǂɔipe}{\dictsense{adv.}{maybe.not}{}}
\dictentry{ǂooṭri}{ǂɔːʈ͡ʂi}{\dictsense{n.}{mountain}{}}
\dictentry{ǂoĩ}{ǂɔ̃ĩ}{\dictsense{v.\-intr.}{fly}{}}
\dictentry{ǂo̰õ}{ǂɔ̰̃ː}{\dictsense{clf.}{lids, lid-like objects, covers, doors}{}}
\dictentry{ǂu}{ǂu}{\dictsense{conj.}{and}{}}
\dictentry{ǂum}{ǂmː}{\dictsense{n.}{canteen}{}}
\dictentry{ǂṵm}{ǂm̰ː}{\dictsense{clf.}{liquids, drops, rain, beverages}{}}
\dictentry{ǂṵʼu}{ǂṵʔu}{\dictsense{n.}{the sun, sunlight, daylight, the daytime sky}{\dictclassifiers{\dictref{aje}}}}
\lettersection{ǂʼ}\dictentry{aǂʼui}{aǂˀui}{\dictsense{clf.}{wooden}{}}
\dictentry{ǂʼa}{ǂˀa}{\dictsense{clf.}{larger herbivores, elands, elephants, giraffes, etc}{}}
\lettersection{ǂx}\dictentry{ǂxa̰a}{ǂ͡χa̰ː}{\dictsense{v.\-tr.}{hit, strike with a loud sound}{}\dictsensesep\dictsense{v.\-tr.}{damage, hurt, offend}{}}
\dictentry{ǂxa̰uʼa}{ǂ͡χa̰ṵʔa}{\dictsense{n.}{effort, strain, force}{}}
\dictentry{ǂxoiṭa}{ǂ͡χɔiʈa}{\dictsense{adj.}{strange}{}}
\dictentry{ǂxoõ}{ǂ͡χɔ̃ː}{\dictsense{n.}{young child}{\dictclassifiers{\dictref{ji}}}}
\dictentry{ǂxoĩ}{ǂ͡χɔ̃ĩ}{\dictsense{post.}{through}{}\dictsensesep\dictsense{post.}{across}{}}
\lettersection{ǂqʼ}\dictentry{ǂqʼaĩ}{ǂ͡qʼãĩ}{\dictsense{v.}{know}{}}
\dictentry{ǂqʼula}{ǂ͡qʼula}{\dictsense{adv.}{before}{}\dictsensesep\dictsense{n.}{(anatomy) back, spine, buttocks}{}}
\lettersection{ɴǂ}\dictentry{aɴǂa̰i}{ãᵑǂa̰ḭ}{\dictsense{adj.}{every}{}}
\dictentry{uɴǂa̰aki}{mᵑǂa̰ːki}{\dictsense{v.\-intr.}{(S) climb}{}}
\dictentry{uɴǂoi}{mᵑǂɔi}{\dictsense{n.}{language, way of speaking}{\dictclassifiers{\dictref{ṇṵĩ}}}}
\dictentry{uɴǂṵu}{mᵑǂṵː}{\dictsense{rel.\-pr.}{which.ERG}{}}
\dictentry{ɴǂaã}{ᵑǂãː}{\dictsense{v.}{complain}{}}
\dictentry{ɴǂoa}{ᵑǂɔa}{\dictsense{v.\-intr.}{discover}{}}
\dictentry{ɴǂoichi}{ᵑǂɔit͡ʃi}{\dictsense{card.\-num.}{seven}{}}
\dictentry{ɴǂo̰õ}{ᵑǂɔ̰̃ː}{\dictsense{n.}{bed}{}}
\dictentry{ɴǂuĩ}{ᵑǂmĩ}{\dictsense{v.}{kick}{}}
\dictentry{ɴǂṵu}{ᵑǂṵː}{\dictsense{v.\-intr.}{die}{}\dictsensesep\dictsense{n.}{death, especially personified as "Lady Death"}{\dictclassifiers{\dictref{mḛʼe}}}}
\lettersection{ɴǂʼ}\dictentry{ɴǂʼa}{ᵑǂˀa}{\dictsense{v.\-tr.}{have sex, mate, copulate; (P) is usually the woman or passive role, and (A) or (T) is the male or active.}{}}
\dictentry{ɴǂʼi}{ᵑǂˀi}{\dictsense{Particle}{NEG, not, negates a preceding adverb, adjective or noun (not verbs)}{}}
\dictentry{ɴǂʼɴǂʼa}{ᵑǂˀᵑǂˀa}{\dictsense{v.\-intr.}{(of multiple people) have sex with eachother}{}}
\lettersection{ɴǂx}\dictentry{ɴǂxa}{ᵑǂ͡ʁa}{\dictsense{adv.}{always, often, it is a regular occurrence that}{}}
\lettersection{ṣǂ}\dictentry{ṣǂa}{ʂǂa}{\dictsense{n.}{human being, person}{}}
\dictentry{ṣǂum}{ʂǂmː}{\dictsense{n.}{knife}{\dictclassifiers{\dictref{ǃxa}}}}
\lettersection{ṣǂʼ}\dictentry{ṣǂʼo}{ʂǂˀɔ}{\dictsense{v.\-intr.}{sweat}{}\dictsensesep\dictsense{n.}{sweat}{\dictclassifiers{\dictref{ǂṵm}}}}
\lettersection{ṣǂqʼ}\dictentry{ṣǂqʼo}{ʂǂ͡qʼɔ}{\dictsense{n.}{neck}{}}
\lettersection{ǃ}\dictentry{aǃṵuma}{aǃṵːma}{\dictsense{n.}{hand}{}\dictsensesep\dictsense{card.\-num.}{ten}{}}
\dictentry{eǃum}{eǃmː}{\dictsense{pers.\-pr.}{me (ACC), to me}{}}
\dictentry{eǃumṭa}{eǃmːʈa}{\dictsense{pers.\-pr.}{1.PL.EXCL.ACC}{}}
\dictentry{iǃo̰orri}{iǃɔ̰ːrri}{\dictsense{v.}{eat}{}}
\dictentry{oǃao}{ɔǃaɔ}{\dictsense{adj.}{old}{}}
\dictentry{uǃa̰ama}{uǃa̰ːma}{\dictsense{n.}{misstep, mistake}{}}
\dictentry{uǃoõ}{uǃɔ̃ː}{\dictsense{n.}{year}{}}
\dictentry{uǃo̰i}{uǃɔ̰ḭ}{\dictsense{adj.}{jittery, irritable, violent, uneasy, startled}{\dictexample{Uǃʼui ʼa uǃòi ṇàã ǂʼa!}{You startled the eland!}}}
\dictentry{uǃuuji}{uǃuːɟi}{\dictsense{adv.}{for the duration / over the course of a whole day, starting in the morning and ending in the evening, possibly repeating over multiple days each occupied completely by the action}{}\dictsensesep\dictsense{n.}{day, span of a day, time duration between sunrise and sunset}{\dictclassifiers{\dictref{ʼurri}}}}
\dictentry{ǃa}{ǃa}{\dictsense{pers.\-pr.}{I (ERG), me, from me, because of mes}{}}
\dictentry{ǃaala}{ǃaːla}{\dictsense{post.}{under, below}{}\dictsensesep\dictsense{post.}{moving by means of, travelling by}{}\dictsensesep\dictsense{n.}{palm (of hand), sole (foot)}{}}
\dictentry{ǃam}{ǃãm}{\dictsense{clf.}{spirits, ghosts, ancestors, the dead}{}\dictsensesep\dictsense{clf.}{trees}{}}
\dictentry{ǃauṭa}{ǃauʈa}{\dictsense{pers.\-pr.}{Us, excluding you (ERG)}{}}
\dictentry{ǃoa}{ǃɔa}{\dictsense{adj.}{idiot}{}}
\dictentry{ǃoi}{ǃɔi}{\dictsense{post.}{for the benefit of, for the purpose of giving to}{}\dictsensesep\dictsense{post.}{for the purpose/with the intent of going to, travelling to, or moving towards}{}}
\dictentry{ǃooja}{ǃɔːɟa}{\dictsense{preverb}{Negative Deontic marker -- the speaker wishes, desires, requests, plans or intends for the action not to happen.}{}}
\dictentry{ǃoorro}{ǃɔːrrɔ}{\dictsense{n.}{urine}{\dictclassifiers{\dictref{ǂṵm}}}\dictsensesep\dictsense{v.\-intr.}{urinate}{}}
\dictentry{ǃooṉo}{ǃɔːṉɔ}{\dictsense{n.}{boy}{\dictclassifiers{\dictref{ji},\dictref{ṉui}}}}
\dictentry{ǃope}{ǃɔpe}{\dictsense{v.\-tr.}{kill, slay, murder}{}}
\dictentry{ǃo̰o}{ǃɔ̰ː}{\dictsense{card.\-num.}{one, non-plural}{}\dictsensesep\dictsense{adj.}{lone, alone, unaccompained, unpaired}{}}
\dictentry{ǃo̰otɬo}{ǃɔ̰ːt͡ɬɔ}{\dictsense{n.}{vulva}{}}
\dictentry{ǃumña}{ǃmːɲa}{\dictsense{n.}{rain}{}}
\dictentry{ǃumñi}{ǃmːɲi}{\dictsense{pers.\-pr.}{2.S.ACC}{}}
\dictentry{ǃumṭa}{ǃmːʈa}{\dictsense{pers.\-pr.}{2.PL.ACC}{}}
\dictentry{ǃuuli}{ǃuːli}{\dictsense{n.}{celebration, party}{}}
\dictentry{ǃuũǃoi}{ǃmːǃɔi}{\dictsense{N/A}{hello, hi}{}}
\lettersection{ǃʼ}\dictentry{eǃʼaṇi}{eǃˀaɳi}{\dictsense{v.}{sing}{}}
\dictentry{oǃʼo}{ɔǃˀɔ}{\dictsense{v.\-intr.}{arrive (among others), join (Dat), meet up with others (Dat)}{}}
\dictentry{uǃʼui}{uǃˀui}{\dictsense{pers.\-pr.}{2.ERG}{}}
\dictentry{ǃʼa}{ǃˀa}{\dictsense{post.}{besides, to the side of, near}{}\dictsensesep\dictsense{n.}{hips, iliac crests}{\dictclassifiers{\dictref{ʇṵu}}}}
\dictentry{ǃʼiṉa}{ǃˀiṉa}{\dictsense{n.}{boat}{}}
\dictentry{ǃʼoã}{ǃˀɔ̃ã}{\dictsense{n.}{fingernail}{}\dictsensesep\dictsense{n.}{(of a location or a stretch in time) end, endpoint, boundary, finish, completion, last portion}{}}
\dictentry{ǃʼuulu}{ǃˀuːlu}{\dictsense{v.\-tr.}{bite}{}\dictsensesep\dictsense{v.\-intr.}{feel pain, especially itching of the skin}{}}
\lettersection{ǃx}\dictentry{oǃxo̰oji}{ɔǃ͡χɔ̰ːɟi}{\dictsense{v.}{get stuck, become unable to move or act}{}}
\dictentry{oǃxu}{ɔǃ͡χu}{\dictsense{n.}{walking cane}{}}
\dictentry{uǃxa̰m}{uǃ͡χã̰m̰}{\dictsense{v.}{concern}{}}
\dictentry{ǃxa}{ǃ͡χa}{\dictsense{clf.}{blades, things with a sharp edge or point, arrows, spears, teeth}{}}
\dictentry{ǃxaje}{ǃ͡χaɟe}{\dictsense{v.\-tr.}{open}{}}
\dictentry{ǃxape}{ǃ͡χape}{\dictsense{adj.}{happy}{}}
\dictentry{ǃxa̰a}{ǃ͡χa̰ː}{\dictsense{n.}{house, hut}{}\dictsensesep\dictsense{n.}{roof}{\dictclassifiers{\dictref{ǂo̰õ}}}}
\dictentry{ǃxo̰o}{ǃ͡χɔ̰ː}{\dictsense{pers.\-pr.}{you and I}{}}
\lettersection{ǃqʼ}\dictentry{uǃqʼa}{uǃ͡qʼa}{\dictsense{n.}{wind}{\dictclassifiers{\dictref{maã}}}}
\dictentry{uǃqʼo}{uǃ͡qʼɔ}{\dictsense{v.\-tr.}{peel, scrape or remove a covering, protective layer, film, piece of clothing}{}}
\dictentry{ǃqʼao}{ǃ͡qʼaɔ}{\dictsense{n.}{clock}{}}
\dictentry{ǃqʼaṯi}{ǃ͡qʼat͡s̪i}{\dictsense{refl.\-pr.}{self.ACC, placed in P position, marks that (C) is also acting as (P).}{}}
\dictentry{ǃqʼoa}{ǃ͡qʼɔa}{\dictsense{v.\-tr.}{steal}{}}
\lettersection{ǃʛ}\dictentry{ǃʛoa}{ǃ͡ʛɔa}{\dictvariantof{\dictref{ǃoa}}}
\lettersection{ɴǃ}\dictentry{ɴǃai}{ᵑǃai}{\dictsense{adj.}{similar to, akin to}{}}
\dictentry{ɴǃali}{ᵑǃali}{\dictsense{n.}{father}{\dictclassifiers{\dictref{ñḛʼe}}}}
\dictentry{ɴǃa̰a}{ᵑǃa̰ː}{\dictsense{n.}{fire, firepit}{}\dictsensesep\dictsense{v.\-tr.}{cook (on fire), roast}{}}
\dictentry{ɴǃooṭo}{ᵑǃɔːʈɔ}{\dictsense{n.}{chicken}{\dictclassifiers{\dictref{uṭu}}}}
\dictentry{ɴǃo̰õ}{ᵑǃɔ̰̃ː}{\dictsense{n.}{North}{}}
\dictentry{ɴǃo̰ʼo}{ᵑǃɔ̰ʔɔ}{\dictsense{v.\-intr.}{breathe}{}}
\dictentry{ɴǃui}{ᵑǃui}{\dictsense{clf.}{repeated.actions}{}}
\lettersection{ɴǃʼ}\dictentry{ɴǃʼoirre}{ᵑǃˀɔirre}{\dictsense{v.}{decide}{}}
\lettersection{ɴǃx}\dictentry{aɴǃxo̰o}{ãᵑǃ͡ʁɔ̰ː}{\dictsense{pers.\-pr.}{me and you}{}}
\dictentry{ɴǃxa̰a}{ᵑǃ͡ʁa̰ː}{\dictsense{post.}{inside}{}}
\dictentry{ɴǃxṵu}{ᵑǃ͡ʁṵː}{\dictsense{adv.}{more}{}}
\lettersection{šǃ}\dictentry{išǃuka}{iʃǃuka}{\dictsense{N/A}{the very same}{}}
\dictentry{ušǃuupa}{uʃǃuːpa}{\dictsense{v.}{re-organize}{}}
\dictentry{šǃaʼa}{ʃǃaʔa}{\dictsense{v.\-tr.}{memorize, either by practice or through an intense experience resulting in a vivid memory}{}}
\dictentry{šǃo}{ʃǃɔ}{\dictsense{Copulative verb}{be temporarily, be contingentially, there be, be present}{}}
\dictentry{šǃo ... iñi}{ʃǃɔ ... iɲi}{\dictsense{v.}{(smth) be over, be on top of}{}\dictsensesep\dictsense{v.}{(actions \& events) be involved in, act in, perform, be busy with}{}\dictsensesep\dictsense{v.}{lie on, lay down on, cover}{\dictexample{šǃo ku nǂòõ iňi}{he is lying on the bed}}}
\dictentry{šǃoiñe}{ʃǃɔiɲe}{\dictsense{n.}{meat}{\dictclassifiers{\dictref{la̰a}}}}
\dictentry{šǃu}{ʃǃu}{\dictsense{Resumptive marker}{RES}{}}
\lettersection{šǃʼ}\dictentry{šǃʼa}{ʃǃˀa}{\dictsense{n.}{tooth}{\dictclassifiers{\dictref{ʇṵu},\dictref{ǃxa}}}}
\lettersection{šǃqʼ}\dictentry{šǃqʼoi}{ʃǃ͡qʼɔi}{\dictsense{card.\-num.}{five}{}}
\lettersection{ʇ}\dictentry{iʇaã}{iǀãː}{\dictsense{v.\-intr.}{do nothing, be slacking, loiter}{\dictexample{Nǂxa iʇaã ǃXaoʼaã ku.}{ǃXaoʼaã is always slacking.}}\dictsensesep\dictsense{v.\-intr.}{lie down, be on the ground}{}\dictsensesep\dictsense{v.\-tr.}{(a dead person) bury}{}}
\dictentry{uʇum}{uǀmː}{\dictsense{v.\-intr.}{stand up, stand up straight, get up}{}\dictsensesep\dictsense{v.\-intr.}{(in SVC) be about to ..., will ..., be likely to ..., plan to ..., intend to ...}{}}
\dictentry{ʇa}{ǀa}{\dictsense{clf.}{feelings, emotions, relationships, mental states}{}}
\dictentry{ʇaala}{ǀaːla}{\dictsense{adv.}{easily}{}}
\dictentry{ʇaula}{ǀaula}{\dictsense{adv.}{once.upon.a.time}{}}
\dictentry{ʇaõ}{ǀãɔ̃}{\dictsense{clf.}{orifices, bodily holes, openings, wounds}{}}
\dictentry{ʇoã}{ǀɔ̃ã}{\dictsense{Ditransitive verb}{(A) gift (T) to (P)}{}}
\dictentry{ʇuli}{ǀuli}{\dictsense{n.}{breast}{\dictclassifiers{\dictref{ʇṵu}}}\dictsensesep\dictsense{n.}{mother}{\dictclassifiers{\dictref{tɬa}}}}
\dictentry{ʇṵu}{ǀṵː}{\dictsense{clf.}{body parts}{}}
\dictentry{ʇṵupa}{ǀṵːpa}{\dictsense{pers.\-pr.}{1.PL.INCL.ERG}{}}
\lettersection{ʇʼ}\dictentry{iʇʼali}{iǀˀali}{\dictsense{n.}{barrier}{}}
\dictentry{iʇʼi}{iǀˀi}{\dictsense{clf.}{small or medium-sized animals, especially smaller mammals}{}}
\dictentry{uʇʼule}{uǀˀule}{\dictsense{v.}{create, make, build, manufacture}{}}
\lettersection{ʇx}\dictentry{ʇxoi}{ǀ͡χɔi}{\dictsense{n.}{cloak}{}}
\lettersection{ʇqʼ}\dictentry{ʇqʼa}{ǀ͡qʼa}{\dictsense{n.}{night}{}}
\dictentry{ʇqʼaãṭa}{ǀ͡qʼãːʈa}{\dictsense{adv.}{at.nighttime}{}}
\lettersection{ɴʇ}\dictentry{uɴʇaã}{mᵑǀãː}{\dictsense{n.}{wolf}{\dictclassifiers{\dictref{ṭa̰a}}}\dictsensesep\dictsense{adj.}{(of a person) unpredictably aggressive, pugnacious, cruel, dangerous}{}}
\dictentry{uɴʇu}{mᵑǀu}{\dictsense{adv.}{truly}{}}
\dictentry{ɴʇa}{ᵑǀa}{\dictsense{n.}{child (as in progeny, son or daughter)}{}}
\dictentry{ɴʇaaṯi}{ᵑǀaːt͡s̪i}{\dictsense{card.\-num.}{nine}{}}
\dictentry{ɴʇai}{ᵑǀai}{\dictsense{post.}{face}{}}
\dictentry{ɴʇa̰ʼã}{ᵑǀã̰ʔã}{\dictsense{n.}{womb, vagina}{\dictclassifiers{\dictref{ʇṵu}}}}
\dictentry{ɴʇum}{ᵑǀmː}{\dictsense{card.\-num.}{two}{}}
\lettersection{ɴʇʼ}\dictentry{oɴʇʼa}{ɔ̃ᵑǀˀa}{\dictsense{v.}{envelop}{}}
\lettersection{sʇ}\dictentry{sʇau}{s̪ǀau}{\dictsense{adj.}{strong}{}}
\dictentry{sʇooṉa}{s̪ǀɔːṉa}{\dictsense{v.\-tr.}{poison.tip}{}}
\dictentry{sʇui}{s̪ǀui}{\dictsense{n.}{snake}{}}
\lettersection{sʇʼ}\dictentry{sʇʼe}{s̪ǀˀe}{\dictsense{card.\-num.}{four}{}}
\dictentry{sʇʼi}{s̪ǀˀi}{\dictsense{clf.}{slender}{}}
\lettersection{sʇqʼ}\dictentry{isʇqʼa}{is̪ǀ͡qʼa}{\dictsense{v.\-tr.}{skin.for.meat}{}}
\lettersection{ʘ}\dictentry{ʘa̰a}{ʘa̰ː}{\dictsense{clf.}{songs, music, singing, sung phrases, rhythmic noises, birdsong}{}\dictsensesep\dictsense{clf.}{dances, dancing, dance moves, rhythmic movements}{}}
\dictentry{ʘui}{ʘui}{\dictsense{v.\-tr.}{want}{}}
\lettersection{ɴʘ}\dictentry{ɴʘaã}{ᵑʘãː}{\dictsense{n.}{air}{}}
\dictentry{ɴʘṵm}{ᵑʘm̰ː}{\dictsense{Ditransitive verb}{(A) suck (substance (T)) from (P), (A) suck on (P)}{}\dictsensesep\dictsense{v.\-tr.}{suckle on, (A) be breastfed by (P)}{\dictexample{ɴʘùm Qʼoaʇùu ǁàa}{Young Qʼoaʇùu is breastfeeding (her child).}}\dictsensesep\dictsense{v.\-tr.}{kiss (on the lips or otherwise)}{}}
\lettersection{ch}\dictentry{cha}{t͡ʃa}{\dictsense{conj.}{while}{}}
\dictentry{chi}{t͡ʃi}{\dictsense{clf.}{aquatic.animals}{}}
\dictentry{chḛʼe}{t͡ʃḛʔe}{\dictsense{n.}{(of tree) trunk, log, column}{}}
\dictentry{chḭichi}{t͡ʃḭːt͡ʃi}{\dictsense{v.}{shine}{}}
\dictentry{uchi}{ut͡ʃi}{\dictsense{pers.\-pr.}{2.INTR}{}}
\lettersection{cʼ}\dictentry{cʼaã}{cʼãː}{\dictsense{n.}{arm}{\dictclassifiers{\dictref{ʇṵu}}}\dictsensesep\dictsense{n.}{wing (of bird)}{\dictclassifiers{\dictref{ʇṵu}}}}
\dictentry{cʼi}{cʼi}{\dictsense{adj.}{such, similar, of the same kind, of the type being discussed}{}}
\lettersection{j}\dictentry{aje}{aɟe}{\dictsense{clf.}{objects and phenomena in the sky, stars, the sun, the moon, comets, clouds, rainbows, sunrises and sunsets, eclipses, etc.}{}}
\dictentry{ja}{ɟa}{\dictsense{pron.}{1S.INTR}{}}
\dictentry{jaṭa}{ɟaʈa}{\dictsense{pers.\-pr.}{1.PL.EXCL.INTR}{}}
\dictentry{ji}{ɟi}{\dictsense{clf.}{children, young people}{}}
\dictentry{jipa}{ɟipa}{\dictsense{n.}{hare}{}}
\dictentry{jḛeñi}{ɟḛːɲi}{\dictsense{N/A}{this}{}}
\dictentry{jṵm}{ɟm̰ː}{\dictsense{v.\-tr.}{surround, circle, flank, in either threatening or protective manner}{}\dictsensesep\dictsense{n.}{circular formation of people or objects, ring of items, a group encircling a centre}{\dictclassifiers{\dictref{ɬǁoi}}}\dictsensesep\dictsense{n.}{group of people by a campfire}{}}
\dictentry{uji}{uɟi}{\dictsense{preverb}{ANTIP}{}}
\lettersection{k}\dictentry{ke}{ke}{\dictsense{clf.}{containers.of.liquids}{}}
\dictentry{ku}{ku}{\dictsense{clf.}{male adults, men}{}}
\dictentry{kuñe}{kuɲe}{\dictsense{adv.}{simply}{}}
\lettersection{l}\dictentry{aliʼa}{aliʔa}{\dictsense{adv.}{merely}{}}
\dictentry{laã}{lãː}{\dictsense{n.}{tongue}{\dictclassifiers{\dictref{ʇṵu}}}}
\dictentry{la̰a}{la̰ː}{\dictsense{clf.}{foul-smelling objects, rotting matter, feces, waste, corpses, infections, pus, diseases, the severely diseased}{}}
\dictentry{la̰ʼã}{lã̰ʔã}{\dictsense{v.\-tr.}{disgust, repulse, cause nausea to, cause to vomit, disturb, distress}{}\dictsensesep\dictsense{v.\-intr.}{be disgusted, nauseous, sick}{}}
\dictentry{loõṇi}{lɔ̃ːɳi}{\dictsense{adj.}{warm, warming}{}\dictsensesep\dictsense{n.}{warmth}{}\dictsensesep\dictsense{adj.}{sensual, seductive, comforting}{}}
\lettersection{m}\dictentry{mau}{mau}{\dictsense{v.\-intr.}{talk}{}\dictsensesep\dictsense{v.\-intr.}{(O) act like (A), makes decision or behaves according to what is expected of (A)}{}\dictsensesep\dictsense{v.\-intr.}{(following in SVC) pretend to ..., deceive oneself or others that you are ...}{}}
\dictentry{maã}{mãː}{\dictsense{clf.}{weather phenomena, winds, rains, sandstorms}{}}
\dictentry{moõ}{mɔ̃ː}{\dictsense{Interrogative pro-form}{what? which?}{}}
\dictentry{moõ ǃoi}{mɔ̃ː ǃɔi}{\dictsense{Interrogative pro-form}{Why? To which end or purpose?}{}}
\dictentry{mḛʼe}{mḛʔe}{\dictsense{clf.}{young childless women, girls who have started menstruating but are yet to be married}{}}
\dictentry{uma}{uma}{\dictsense{pers.\-pr.}{1.PL.INCL.ABS}{}}
\lettersection{p}\dictentry{paala}{paːla}{\dictsense{n.}{wail}{}}
\dictentry{pau}{pau}{\dictsense{adj.}{abundant}{}}
\dictentry{po}{pɔ}{\dictsense{n.}{lips (of the mouth)}{\dictclassifiers{\dictref{ʇṵu}}}\dictsensesep\dictsense{n.}{mouth, oral cavity}{\dictclassifiers{\dictref{ʇaõ}}}}
\dictentry{pṵuṭri}{pṵːʈ͡ʂi}{\dictsense{n.}{mongoose}{}}
\dictentry{upa}{upa}{\dictsense{preverb}{SUBJ}{}}
\lettersection{qʼ}\dictentry{qʼañi}{qʼaɲi}{\dictsense{n.}{vagabond}{}}
\lettersection{r}\dictentry{ra}{ra}{\dictsense{post.}{INSTR}{}}
\lettersection{tɬ}\dictentry{tɬa}{t͡ɬa}{\dictsense{clf.}{adult women}{}}
\dictentry{tɬa̰ʼa}{t͡ɬa̰ʔa}{\dictsense{n.}{fish}{}}
\dictentry{tɬḭi}{t͡ɬḭː}{\dictsense{preverb}{COND}{}}
\lettersection{tɬʼ}\dictentry{tɬʼoi}{t͡ɬʼɔi}{\dictsense{v.\-intr.}{leave, abandon (ABL), exit}{}\dictsensesep\dictsense{v.\-intr.}{(in SVC) stop, interrupt ..., finish, conclude ...}{}}
\dictentry{utɬʼe}{ut͡ɬʼe}{\dictsense{n.}{walking path, paved path, dirt road}{}\dictsensesep\dictsense{n.}{groove, incision, indented strip}{}}
\lettersection{ñ}\dictentry{eña}{eɲa}{\dictsense{v.\-tr.}{surrender (smth.), let go of, unwillingly offer}{}}
\dictentry{eña ṣǂqʼo}{eɲa ʂǂ͡qʼɔ}{\dictsense{v.\-tr.}{(A) surrender oneself, give up (lit. offer neck)}{}}
\dictentry{iñi}{iɲi}{\dictsense{post.}{over, on top of, above}{}\dictsensesep\dictsense{Subordinating connective}{provided that, resting on the fact that, the fact that ... guarantees that ...}{}\dictsensesep\dictsense{n.}{head}{}}
\dictentry{ña̰ã}{ɲã̰ː}{\dictsense{N/A}{liver}{}}
\dictentry{ña̰ʼaña}{ɲa̰ʔaɲa}{\dictsense{n.}{hunter}{}}
\dictentry{ñiĩ}{ɲĩː}{\dictsense{adv.}{instead, yes instead, but ... yes; marks that the clause holds despite or in opposition to an example previously given.}{\dictexample{Uǃuuji ʼmʼm ɴǁoi ǁo̰i ǁqʼoṯe, ṇam ṣǂa ñiĩ.}{Prey can't run for a whole day, but a human can.}}}
\dictentry{ñḛʼe}{ɲḛʔe}{\dictsense{clf.}{elder men}{}}
\lettersection{ŋ}\dictentry{uŋa}{uŋa}{\dictsense{adj.}{none}{}}
\dictentry{ŋa̰ã}{ŋã̰ː}{\dictsense{n.}{woman}{}}
\dictentry{ŋum}{ŋmː}{\dictsense{card.\-num.}{twelve}{}}
\dictentry{ŋḛe}{ŋḛː}{\dictsense{n.}{evening, time of sunset}{\dictclassifiers{\dictref{ʼurri}}}\dictsensesep\dictsense{n.}{sunset (the process of sun setting)}{}}
\lettersection{ʼ}\dictentry{ʼa}{ʔa}{\dictsense{post.}{ABL, coming from, originating from, created by, moving away from}{}\dictsensesep\dictsense{post.}{causative marker, denotes causer of the action}{}\dictsensesep\dictsense{post.}{emphatic ergative marker}{}}
\dictentry{ʼa ... ɴǁoi}{ʔa ... ᵑǁɔi}{\dictsense{v.\-tr.}{allow, permit, facilitate, (A) allow (O) to ... (lit. because of (A), (O) can ...)}{}}
\dictentry{ʼai}{ʔai}{\dictsense{adv.}{and (for clauses)}{}}
\dictentry{ʼai ʼai}{ʔai ʔai}{\dictsense{conj.}{and thus, and as a consequence, and immediately after}{}}
\dictentry{ʼa̰o}{ʔa̰ɔ̰}{\dictsense{n.}{water}{\dictclassifiers{\dictref{ǂṵm}}}\dictsensesep\dictsense{v.\-tr.}{fill up with water}{}}
\dictentry{ʼmʼm}{ʔm̰ʔm}{\dictsense{v.\-intr.}{run}{}}
\dictentry{ʼu}{ʔu}{\dictsense{post.}{of}{}}
\dictentry{ʼurri}{ʔurri}{\dictsense{clf.}{timespans, events in time, occurrences, dates, appointments}{}}
\dictentry{ʼutɬa}{ʔut͡ɬa}{\dictsense{n.}{playing ball}{\dictclassifiers{\dictref{ṉoõ}}}}
\dictentry{ʼḛe}{ʔḛː}{\dictsense{post.}{PTV}{}}
\dictentry{ʼḭi}{ʔḭː}{\dictsense{clf.}{slithering.animals}{}}
\dictentry{ʼṵa}{ʔṵa̰}{\dictsense{v.\-tr.}{(someone) meet, make acquaintance of, get to know, greet, receive}{}}
\lettersection{ṇ}\dictentry{ṇam}{ɳãm}{\dictsense{conj.}{but}{}}
\dictentry{ṇam ... ñiĩ}{ɳãm ... ɲĩː}{\dictvariantof{\dictref{ñiĩ}}}
\dictentry{ṇa̰ã}{ɳã̰ː}{\dictsense{n.}{eland}{\dictclassifiers{\dictref{ǂʼa}}}}
\dictentry{ṇe}{ɳe}{\dictsense{preverb}{Positive DEO marker -- the speaker wishes, requests, desires or intends for the action to happen.}{}}
\dictentry{ṇṵĩ}{ɳm̰ḭ̃}{\dictsense{clf.}{spoken word, utterances, phrases, languages, words, voices, thoughts, reasonings, stories}{}}
\lettersection{ṉ}\dictentry{ṉa}{ṉa}{\dictsense{n.}{grass}{}\dictsensesep\dictsense{n.}{hair, head hair, facial hair, body hair}{}\dictsensesep\dictsense{n.}{(of animal) fur}{}}
\dictentry{ṉa̰ã}{ṉã̰ː}{\dictsense{v.}{laugh}{}}
\dictentry{ṉoõ}{ṉɔ̃ː}{\dictsense{clf.}{round tools, round instruments, artificial balls, spheres, globes, round toys}{}}
\dictentry{ṉoṯi}{ṉɔt͡s̪i}{\dictsense{card.\-num.}{eight}{}}
\dictentry{ṉui}{ṉui}{\dictsense{clf.}{persons, people, humans, personified entities, individuals, animate}{}}
\lettersection{ṭ}\dictentry{uṭu}{uʈu}{\dictsense{clf.}{bird}{}}
\dictentry{ṭa̰a}{ʈa̰ː}{\dictsense{clf.}{predatory animals, carnivores}{}}
\dictentry{ṭuma}{ʈuma}{\dictsense{adj.}{great, awesome}{}}
\dictentry{ṭurra}{ʈurra}{\dictsense{adv.}{suddenly}{}}
\dictentry{ṭḛe}{ʈḛː}{\dictsense{clf.}{articles of clothing, cloth, shoes}{}}
\lettersection{ṭr}\dictentry{ṭre}{ʈ͡ʂe}{\dictsense{n.}{peak}{}}
\dictentry{ṭrui}{ʈ͡ʂui}{\dictsense{n.}{nose}{}}
\lettersection{ṭʼ}\dictentry{ṭʼaṇi}{ʈʼaɳi}{\dictsense{v.\-tr.}{remind}{}}
\dictentry{ṭʼoa}{ʈʼɔa}{\dictsense{n.}{corpse, cadaver, carcass (human or animal)}{\dictclassifiers{\dictref{la̰a}}}}
\dictentry{ṭʼṭʼaṇi}{ʈʼʈʼaɳi}{\dictsense{v.\-tr.}{recognize}{}}
\lettersection{ṯ}\dictentry{eṯe}{et͡s̪e}{\dictsense{N/A}{when}{}}
\dictentry{ṯui}{t͡s̪ui}{\dictsense{v.\-intr.}{cry}{}}
\dictentry{ṯuña}{t͡s̪uɲa}{\dictsense{n.}{arrow}{}}
\dictentry{ṯṵu}{t͡s̪ṵː}{\dictsense{v.\-tr.}{throw, launch}{}\dictsensesep\dictsense{v.\-tr.}{produce, spit out, blow, excrete}{}}